Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning

### 2022-08-11 manuscript/manuscript.Rmd
updated figures

```{r fig.cap="Forecast scenarios. A. The Hopf bifurcation: a stable node develops into a limit cycle. B. The saddle-node bifurcation. C. The stochastic transition."}
hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")
panelA <- hopf |> filter(type %in% c("historical", "true"), i < 10) |>
  rename(species = variable) |> 
  mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
  ggplot(aes(t, value, group=interaction(species, i, type), col=species)) + 
  geom_line(show.legend = FALSE, alpha = 0.7) + theme_bw()+  scale_color_grey() +
  ggtitle("A. Hopf bifurcation")

sn <- read_csv("../greta/forecasts/simulation=saddle/saddlenode.csv.gz")
panelB <- sn |> filter(type == "historical", i == 1) |>
  bind_rows(filter(sn, type=="true")|>mutate(type="predicted")) |>
  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
  geom_path(alpha=0.5, show.legend = FALSE) +
  scale_color_grey(end=0.6) + theme_bw() + 
  ggtitle("B. Saddle-node bifurcation")

stoch <- read_csv("../greta/forecasts/simulation=stochastic/stochastic.csv.gz")
panelC <- stoch |> filter(type == "historical", i == 1) |>
  bind_rows(filter(stoch, type=="true", i < 50) |> mutate(type="predicted")) |>
  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
  geom_path(alpha=0.5) + scale_color_grey(end=0.6) + theme_bw() + 
  ggtitle("C. Stochastic transition")

@@

```{r}
#| fig.cap="Overall distribution of skill scores across models, including an ensemble of methods.
#|  Smaller scores are better (indicating smaller errors).
#|  Black bars indicate means, dots indicate all individual predictions over time and replicate 'true'
#| simulations of the given scenario."
shift |> 
  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
  geom_jitter(alpha = 0.06, show.legend = FALSE, shape=".") + 
  stat_summary(fun = mean, shape="|", show.legend = FALSE, col="black") +
  coord_flip() +
  facet_grid(simulation ~ metric, scales = "free") + 
  scale_y_log10() +
  theme_bw() +
  theme(axis.title.y = element_blank())
```

@@

```{r}
forecasts <- read_parquet("data/forecasts.parquet") |>
  filter(forecasting_model != "ml_ensemble")
observations <- read_parquet("data/observations.parquet")

@@

forecasts  |> 
  filter(iter %in% 1:15, simulation == "hopf", variable !="X") |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.1, data = hopf_obs) +
  geom_line(alpha=0.1) +
  geom_line(data = filter(hopf_obs, type=="historical")) +
  facet_grid(forecasting_model ~ variable, scales = "free") +
  theme_bw()  + 
  coord_cartesian(ylim = c(0,15), xlim=c(75, 200)) + 
  ggtitle("Hopf bifurcation")
```

@@

forecasts |> 
  filter(iter %in% 1:15, simulation == "saddle") |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.1, data = saddle_obs) +
  geom_line(alpha=0.1) +
  geom_line(data = filter(saddle_obs, type=="historical")) +
  facet_wrap(~forecasting_model, scales = "free") +
  theme_bw()  + coord_cartesian(ylim = c(0,1)) + 
  ggtitle("Saddle-Node bifurcation")
```

@@

```{r}
#| fig.cap = "Forecasts of the stochastic transition under each model, compared to 15 realizations of the true model. 
#|  In contrast to the other challenges, this case considers the prediction of replicate systems starting from the same initial condition, rather than forecasting the future evolution of the model after the stochastic transition has already occurred."
forecasts |>
  filter(iter %in% 1:15, simulation == "stochastic") |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="observed")) +
  geom_line(alpha=0.2) +
  geom_line(data = filter(stochastic_obs, type=="historical")) +
  facet_wrap(~forecasting_model, scales = "free") +
  theme_bw()  + coord_cartesian(ylim = c(0,0.8)) + 
  ggtitle("Stochastic transition")
```

@@

Examining patterns in the scores over time (Fig 6-7) provides a more nuanced understanding of the forecast dynamics than aggregate scores alone (Fig 2).
In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast,
while MCMC scores are worst in mid-range.
Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation. 
CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast or the other state. 
Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g. by the MCMC predictions. 
The forecast-skill-over time plots illustrate diferent reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e. after the catastrophic transition) are poor.
In the case of the stochastic transition model, the two modes are not structured by horizon but by replicate, with some replicates having transitioned and others still in the original state.

```{r}
#| fig.cap="CRPS scores over time for each scenario.
#|  Each line represents the scores against a replicate of time series observations from 'true model'. 
#|  Patterns show forecast skill generally get worse over time -- 
#| gradually in the case of Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
#|  The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window. 
#|  In the stochastic transition context, this creates two branches, where high values indicate periods of 
#| time when the forecast predicts the wrong state and lower branch indicates the correct state."

>>>>

```{r fig.cap="Forecast scenarios. A. The Hopf bifurcation: a stable node develops into a limit cycle which gradually grows larger in this predator-prey model. B. The saddle-node bifurcation in a single species. C. The stochastic transition in a single species.  Plots show historical data used to train the algorithm in purple, and ensemble replicate simulations of the the future dynamics in yellow. Note how the characteristic time for the critical transition varies across the transitions.", fig.height=6}
hopf <- read_csv("../greta/forecasts/simulation=hopf/hopf_increasing.csv.gz")
panelA <- hopf |> 
  mutate(type = as.factor(type)) |>
  filter(type %in% c("historical", "true"), i < 10) |>
  rename(species = variable) |> 
  mutate(species = forcats::fct_recode(species, host = "H", parasitoid = "P")) |>
  bind_rows(tibble(i=1,t=1,type="predicted", variable="H", value=NA)) |> #placeholder
  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
  ggplot(aes(t, value, group=interaction(species, i, type), col=type, lty=species)) + 
  geom_line(show.legend = TRUE, alpha = 0.7) +
  theme_bw() +  
  scale_color_viridis_d(guide="none") +
  ggtitle("A. Hopf bifurcation")+
  theme(legend.position="top")

sn <- read_csv("../greta/forecasts/simulation=saddle/saddlenode.csv.gz")
panelB <- sn |>
  mutate(type = as.factor(type)) |>
  filter(type == "historical", i == 1) |>
  bind_rows( filter(sn, type=="true")) |>
  bind_rows(tibble(i=1,t=1,type="predicted", variable="N", value=NA)) |> #placeholder
  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
  geom_path(alpha=0.5, show.legend = FALSE) +
  scale_color_viridis_d() +
  theme_bw() + 
  ggtitle("B. Saddle-node bifurcation")

stoch <- read_csv("../greta/forecasts/simulation=stochastic/stochastic.csv.gz")
panelC <- stoch |>
  filter(type == "historical", i == 1) |>
  bind_rows(filter(stoch, type=="true", i < 50)) |> 
  bind_rows(tibble(i=1,t=1,type="predicted", variable="N", value=NA)) |> #placeholder
  mutate(type = forcats::fct_recode(type, "future ensemble" = "true", "forecast" = "predicted")) |>
  ggplot(aes(t, value, col = type, group=interaction(i,type), lty=type)) +
  geom_path(alpha=0.5) + 
  scale_color_viridis_d() +
#  scale_color_grey(end=0.6) + 
  theme_bw() + 
  ggtitle("C. Stochastic transition") +
  theme(legend.position="bottom")

@@

```{r}
#| fig.cap="Overall distribution of skill scores across models, including an ensemble of methods.
#|  Smaller scores are better (indicating smaller errors).
#|  Black bars indicate means, dots indicate all individual predictions over time and replicate 'true'
#| simulations of the given scenario."
shift |> 
  ggplot(aes(forecasting_model, score, color=forecasting_model)) + 
  geom_jitter(alpha = 0.06, show.legend = FALSE, shape=".") + 
  stat_summary(fun = mean, shape="|", show.legend = FALSE, col="black") +
  scale_x_discrete(limits = rev) +
  coord_flip() + 
  facet_grid(simulation ~ metric, scales = "free") + 
  scale_y_log10() +
  theme_bw() +
  theme(axis.title.y = element_blank())
```

@@

```{r}
forecasts <- read_parquet("data/forecasts.parquet") |>
  filter(forecasting_model != "ml_ensemble") |> 
  bind_rows(tibble(type= c("future ensemble", "historical"))) |>
  mutate(type = forcats::fct_recode(type, "future ensemble" = "observed", "forecast" = "predicted"))

observations <- read_parquet("data/observations.parquet") |> 
  mutate(type = forcats::fct_recode(type, "future ensemble" = "observed", "forecast" = "predicted"))

@@

forecasts  |> 
  filter(iter %in% 1:15, simulation == "hopf", variable !="X") |> 
  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |>
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.1) +
  geom_line(alpha=0.1, data = hopf_obs) +
  geom_line(data = filter(hopf_obs, type=="historical")) +
  facet_grid(forecasting_model ~ variable, scales = "free") +
  theme_bw()  + 
  scale_color_viridis_d() +
  coord_cartesian(ylim = c(0,15), xlim=c(75, 200)) + 
  ggtitle("Hopf bifurcation")
```

@@

forecasts |> 
  filter(iter %in% 1:15, simulation == "saddle") |> 
  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.1) +
  geom_line(alpha=0.1, data = saddle_obs) +
  geom_line(data = filter(saddle_obs, type=="historical")) +
  facet_wrap(~forecasting_model, scales = "free") +
  theme_bw()  + 
  scale_color_viridis_d() +
  coord_cartesian(ylim = c(0,1)) + 
  ggtitle("Saddle-Node bifurcation")
```

@@

```{r}
#| fig.cap = "Forecasts of the stochastic transition under each model, compared to 15 realizations of the true model. 
#|  In contrast to the other challenges, this case considers the prediction of replicate systems starting from the same initial condition, rather than forecasting the future evolution of the model after the stochastic transition has already occurred."
forecasts |>
  filter(iter %in% 1:15, simulation == "stochastic") |> 
  bind_rows(tibble(type= c("future ensemble", "historical"), forecasting_model="MCMC", variable="host")) |>
  mutate(type = factor(type,levels=c("historical","forecast", "future ensemble"))) |> 
  ggplot(aes(x=t, y=value, group = interaction(iter, variable), col=type)) + 
  geom_line(alpha=0.2) +
  geom_line(alpha=0.2, data = filter(stochastic_obs, type=="future ensemble")) +
  geom_line(data = filter(stochastic_obs, type=="historical")) +
  facet_wrap(~forecasting_model, scales = "free") +
  theme_bw()  + 
  scale_color_viridis_d() +
  coord_cartesian(ylim = c(0,0.8)) + 
  ggtitle("Stochastic transition")
```

@@

Examining patterns in the scores over time (Fig 6-7) provides a more nuanced understanding of the forecast dynamics than aggregate scores alone (Fig 2).
In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast,
while MCMC scores are worst in mid-range.
Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation. 
CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast or the other state. 
Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g. by the MCMC predictions. 
The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Figure 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e. after the catastrophic transition) are poor.
In the case of the stochastic transition model, the two modes are not structured by horizon but by replicate, with some replicates having transitioned and others still in the original state.

```{r}
#| fig.cap="CRPS scores over time for each scenario.
#|  Each line represents the scores against a replicate of time series observations from 'true model'. 
#|  Patterns show forecast skill generally get worse over time -- gradually 
#| in the case of Hopf bifurcation or suddenly in response to the saddle-node bifurcation. 
#|  The MCMC shows a sharp improvement in forecast skill over time as it passes the transition window. 
#|  In the stochastic transition context, this creates two branches, where high values indicate periods of 
#| time when the forecast predicts the wrong state and lower branch indicates the correct state."
